Comparing Generative Pre-trained Transformer Models to Physicians for Assigning Causes of Death from Verbal Autopsy: A Case Study of 6939 Deaths in Sierra Leone from 2019-2022

### 2024-04-14 manuscript/manuscript.tex
Done neo writing edits

This study uses 6939 physician agreed records of 11,920 records collected in 2019 to 2022 from the Healthy Sierra Leone (HEAL-SL) study as described in Section \ref{methods-data}. Section \ref{methods-models} describes the four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, for COD assignment, and their inputs and outputs, while section \ref{methods-eval} details the performance evaluation of the four models relative to physicians overall, by ages, and by COD, using population and individual level metrics. Figure \ref{fig-methods} details the methods used in this study.

@@

Across the adult age ranges, GPT-4 had the highest PCCC values at or above 0.59, followed by GPT-3.5, while InterVA-5 and InSilicoVA had the lowest PCCC values at or below 0.46. GPT-4 PCCC was between 0.61 and 0.72 for young to middle age ranges between 10 to 49 years, and at or below 0.62 for older age ranges between 50 to 69 years (Figure \ref{fig-perf-agerange-adult}). Each age range had between 135 (4\%) to 575 (15\%) records with the 65-69 year age range having the most records (n=575, 15\%), while other age ranges ranged from 4-9\% of all records. Figure \ref{fig-perf-cod-adult} shows the highest and lowest performing models by PCCC across 18 adult CODs. The highest PCCC values were GPT-4 between 0.36 to 0.99 for 10 of 17 CODs (GPT-3.5, InterVA-5, and InSilicoVA within 0.05 for 3, 1, and 1 of these 10 CODs respectively), InSilicoVA for tuberculosis and road and transport injuries at 0.72 and 0.84 (GPT-4 and InterVA-5 within 0.05), and GPT-3.5 between 0.44 and 0.94 for 5 of 17 CODs (GPT-4 within 0.05 for malaria). The lowest PCCC values were InterVA-5 between 0 to 0.79 for 8 of 17 CODs (InSilicoVA and GPT-3.5 within 0.05 for cancers and unspecified infections respectively), InSilicoVA between 0.01 to 0.41 for 6 of 17 CODs (InterVA-5 within 0.05 for other cardiovascular diseases and diarrhoeal diseases), and GPT-3.5 at 0.39 for tuberculosis and 0.54 for road and transport injuries. Chronic respiratory diseases had the largest differences between minimum and maximum PCCC at 0.93. Maternal conditions, and road and transport injuries, had higher PCCC values above or at 0.67 (with the exception of GPT-3.5 for road and transportation injuries as an outlier), where there were small differences between minimum and maximum PCCC below or at 0.2. Ischemic heart disease and unspecified infections had PCCC values below or at 0.65, where there were small differences between minimum and maximum PCCC at or below 0.14. Each adult COD had between 27 (\textless 1\%) to 799 records (21\%) with malaria (n=799, 21\%), other noncommunicable diseases (n=414, 11\%), and unspecified infections (n=389, 10\%) making 42\% of all records, and other CODs between 1-9\% of all records. GPT-4 had the highest PCCC for both male and female records at 0.65 and 0.64 respectively, while all other models had PCCC at or below 0.6 (see Figure \ref{fig-perf-sex-adult} in Appendix \ref{app-perf-misc}).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.7\textwidth]{fig-perf-cod-adult.pdf}
\caption{Model performance for adult records by COD. GPT-3.5, GPT-4, InterVA-5, and InSilicoVA model performance for assigning CGHR-10 CODs in the adult age group (12 to 69 Years) by COD. PCCC values are sorted by the highest to lowest PCCC for each COD. Only models with PCCC values within .05 of the highest and lowest PCCC are shown.}\label{fig-perf-cod-adult}
\end{figure}

@@

Across the child age ranges, the highest PCCC was observed for GPT-4 at 0.59 and 0.66 for 1-5 years and 6-11 years (GPT-3.5 within 0.05), GPT-3.5 at 0.54 for 6-11 months, and InSilicoVA at 0.5 for 1-5 months (Figure \ref{fig-perf-agerange-child}). Each child age range had between 308 (12\%) to 1633 (62\%) records with most records in the 1-5 years range at 1633 records (62\%) and the other age ranges consisting of 12-14\% of all records. For child CODs, the highest PCCC observed was for GPT-4 for 4 of 8 CODs, GPT-3.5 for 3 of 8 CODs (InSilicoVA within 0.05 for other infections), and InSilicoVA for pneumonia (Figure \ref{fig-perf-cod-child}). The lowest PCCC across child CODs was observed for InSilicoVA for epilepsy, leukaemia, and other noncommunicable diseases and malaria, InterVA-5 for pneumonia, nutritional deficiencies, and other infections. GPT-3.5, InSilicoVA, and InterVA-5 all had similar PCCC values greater than 0.8 as the lowest PCCC for injuries. Epilepsy, leukaemia, and other noncommunicable disease had the largest difference between minimum and maximum PCCC at 0.68, while Ill-defined had the next largest difference at 0.67 with InSilicoVA and InterVA-5 at 0 PCCC. Malaria and other infections had low PCCC values below 0.58 across models with small differences between minimum and maximum PCCC at 0.18 and 0.14 respectively. Injuries and diarrhoeal diseases had high PCC values above 0.6 across models with small differences between minimum and maximum PCC at 0.14 and 0.3 respectively. Each child COD had between 11 (\textless 1\%) to 1382 (52\%) records with malaria (n=1382, 52\%), other infections (n=667, 25\%), and diarrhoeal diseases (n=175, 7\%) making 84\% of all records. Nutritional deficiencies had less than 1\% of all records, while other CODs had between 1-3\% of all records. GPT-3.5 had the highest PCCC for both male and female records at 0.56 and 0.58 with GPT-4 within 0.05 PCCC at 0.55 and 0.54 respectively, while all other models had PCCC below 0.5 (see Figure \ref{fig-perf-sex-child} in Appendix \ref{app-perf-misc}).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.7\textwidth]{fig-perf-cod-child.pdf}
\caption{Model performance for child records by COD. GPT-3.5, GPT-4, InterVA-5, and InSilicoVA model performance for assigning CGHR-10 CODs in the child age group (28 days to 11 years) by COD. PCCC values are sorted by the highest to lowest PCCC for each COD. Only models with PCCC values within .05 of the highest and lowest PCCC are shown.}\label{fig-perf-cod-child}
\end{figure}

@@

GPT-4 and InSilicoVA had the highest PCCC for 0-6 day and 7-27 day neonatal age ranges at 0.62 and 0.67 respectively, while InterVA-5 had the lowest PCCC across all neonatal age ranges (GPT-3.5 within 0.05) as seen in Figure \ref{fig-perf-agerange-neo}. There were small differences between minimum and maximum PCCC across all age ranges at 0.14 and 0.12 for 0-6 days and 7-27 days respectively. Most records were in the 0-6 day age range at 395 (83\%) records, while the rest were in the 7-27 day age range at 82 (17\%) records. Figure \ref{fig-perf-cod-neo} shows the performance of models by PCCC across 7 neonatal CODs. GPT-4 had the highest PCCC for 4 of the 7 CODs (GPT-3.5 identical to GPT-4 for congenital anomalies at 1 PCCC). InSilicoVA, GPT-3.5, and InterVA-5 had the highest PCCC for neonatal infections at 0.87, ill-defined at 0.59 (GPT-4 at 0.55), and birth asphyxia and birth trauma at 0.59 respectively. Congenital anomalies and other had less or equal to 5 records with large difference of 1 between minimum and maximum PCCC. Neonatal infections and stillbirth had smaller differences between minimum and maximum PCCC at 0.38 and 0.24 respectively with larger PCCC values at or above 0.49. Birth asphyxia and birth trauma also had smaller differences between minimum and maximum PCCC at 0.24 but had PCCC values at or below 0.59. Prematurity and low birthweight and ill-defined had large differences between minimum and maximum PCCC at 0.62 and 0.58 respectively. Each neonatal COD had between 2 (\textless 1\%) to 172 (38\%) records, with stillbirth (n=172, 36\%), birth asphyxia and birth trauma (n=103, 22\%), and neonatal infections (n=99, 21\%), and prematurity and low birthweight (n=73, 15\%) making 94\% of all records, while the rest are in congenital anomalies, other, and ill-defined CODs. GPT-4 had the highest PCCC for both female and male records at 0.65 and 0.58 (InSilicoVA at 0.56) respectively, while all other models had PCCC at or below 0.56 (see Figure \ref{fig-perf-sex-neo} in Appendix \ref{app-perf-misc}).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.7\textwidth]{fig-perf-cod-neo.pdf}
\caption{Model performance for neonatal records by COD. GPT-3.5, GPT-4, InterVA-5, and InSilicoVA model performance for assigning CGHR-10 CODs in the neonatal age group (under 28 years) by COD. PCCC values are sorted by the highest to lowest PCCC for each COD. Only models with PCCC values within .05 of the highest and lowest PCCC are shown.}\label{fig-perf-cod-neo}
\end{figure}

>>>>

This study uses 6939 physician agreed records of 11,920 records collected in 2019 to 2022 from the Healthy Sierra Leone (HEAL-SL) study as described in Section \ref{methods-data}. Section \ref{methods-models} describes the four models, GPT-3.5, GPT-4, InterVA-5, and InSilicoVA, for COD assignment, and their inputs and outputs, while section \ref{methods-eval} details the performance evaluation of the four models relative to physicians overall, by ages, and by COD, using population and individual level metrics. Figure \ref{fig-methods} in Appendix \ref{app-methods} details the methods used in this study.

@@

Across the adult age ranges, GPT-4 had the highest PCCC values at or above 0.59, followed by GPT-3.5, while InterVA-5 and InSilicoVA had the lowest PCCC values at or below 0.46. GPT-4 PCCC was between 0.61 and 0.72 for young to middle age ranges between 10 to 49 years, and at or below 0.62 for older age ranges between 50 to 69 years (Figure \ref{fig-perf-agerange-adult}). Each age range had between 135 (4\%) to 575 (15\%) records with the 65-69 year age range having the most records (n=575, 15\%), while other age ranges ranged from 4-9\% of all records. Figure \ref{fig-perf-cod-adult} shows the highest and lowest performing models by PCCC across 17 adult CODs excluding suicide due to the less notable 3 (\textless 1\%) records. The highest PCCC values were GPT-4 between 0.36 to 0.99 for 10 of 17 CODs (GPT-3.5, InterVA-5, and InSilicoVA within 0.05 for 3, 1, and 1 of these 10 CODs respectively), InSilicoVA for tuberculosis and road and transport injuries at 0.72 and 0.84 (GPT-4 and InterVA-5 within 0.05), and GPT-3.5 between 0.44 and 0.94 for 5 of 17 CODs (GPT-4 within 0.05 for malaria). The lowest PCCC values were InterVA-5 between 0 to 0.79 for 8 of 17 CODs (InSilicoVA and GPT-3.5 within 0.05 for cancers and unspecified infections respectively), InSilicoVA between 0.01 to 0.41 for 6 of 17 CODs (InterVA-5 within 0.05 for other cardiovascular diseases and diarrhoeal diseases), and GPT-3.5 at 0.39 for tuberculosis and 0.54 for road and transport injuries. Chronic respiratory diseases had the largest differences between minimum and maximum PCCC at 0.93. Maternal conditions, and road and transport injuries, had higher PCCC values above or at 0.67 (with the exception of GPT-3.5 for road and transportation injuries as an outlier), where there were small differences between minimum and maximum PCCC below or at 0.2. Ischemic heart disease and unspecified infections had PCCC values below or at 0.65, where there were small differences between minimum and maximum PCCC at or below 0.14. Each adult COD had between 27 (\textless 1\%) to 799 records (21\%) with malaria (n=799, 21\%), other noncommunicable diseases (n=414, 11\%), and unspecified infections (n=389, 10\%) making 42\% of all records, and other CODs between 1-9\% of all records. GPT-4 had the highest PCCC for both male and female records at 0.65 and 0.64 respectively, while all other models had PCCC at or below 0.6 (see Figure \ref{fig-perf-sex-adult} in Appendix \ref{app-perf-misc}).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.7\textwidth]{fig-perf-cod-adult.pdf}
\caption{Model performance for adult records by COD. GPT-3.5, GPT-4, InterVA-5, and InSilicoVA model performance for assigning CGHR-10 CODs in the adult age group (12 to 69 Years) by COD. PCCC values are sorted by the highest to lowest PCCC for each COD. Only models with PCCC values within .05 of the highest and lowest PCCC are shown. Suicide is not notable due to a small sample size of 3.}\label{fig-perf-cod-adult}
\end{figure}

@@

Across the child age ranges, the highest PCCC was observed for GPT-4 at 0.59 and 0.66 for 1-5 years and 6-11 years (GPT-3.5 within 0.05), GPT-3.5 at 0.54 for 6-11 months, and InSilicoVA at 0.5 for 1-5 months (Figure \ref{fig-perf-agerange-child}). Each child age range had between 308 (12\%) to 1633 (62\%) records with most records in the 1-5 years range at 1633 records (62\%) and the other age ranges consisting of 12-14\% of all records. For the 8 child CODs (excluding congenital anomalies due to small sample size of 1 (\textless 1\%) record), the highest PCCC observed was for GPT-4 for 4 of 8 CODs, GPT-3.5 for 3 of 8 CODs (InSilicoVA within 0.05 for other infections), and InSilicoVA for pneumonia (Figure \ref{fig-perf-cod-child}). The lowest PCCC across the 8 child CODs was observed for InSilicoVA for epilepsy, leukaemia, and other noncommunicable diseases and malaria, InterVA-5 for pneumonia, nutritional deficiencies, and other infections. GPT-3.5, InSilicoVA, and InterVA-5 all had similar PCCC values greater than 0.8 as the lowest PCCC for injuries. Epilepsy, leukaemia, and other noncommunicable disease had the largest difference between minimum and maximum PCCC at 0.68, while Ill-defined had the next largest difference at 0.67 with InSilicoVA and InterVA-5 at 0 PCCC. Malaria and other infections had low PCCC values below 0.58 across models with small differences between minimum and maximum PCCC at 0.18 and 0.14 respectively. Injuries and diarrhoeal diseases had high PCC values above 0.6 across models with small differences between minimum and maximum PCC at 0.14 and 0.3 respectively. Each of the 8 child CODs had between 11 (\textless 1\%) to 1382 (52\%) records with malaria (n=1382, 52\%), other infections (n=667, 25\%), and diarrhoeal diseases (n=175, 7\%) making 84\% of all records. Nutritional deficiencies had less than 1\% of all records, while other CODs had between 1-3\% of all records. GPT-3.5 had the highest PCCC for both male and female records at 0.56 and 0.58 with GPT-4 within 0.05 PCCC at 0.55 and 0.54 respectively, while all other models had PCCC below 0.5 (see Figure \ref{fig-perf-sex-child} in Appendix \ref{app-perf-misc}).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.7\textwidth]{fig-perf-cod-child.pdf}
\caption{Model performance for child records by COD. GPT-3.5, GPT-4, InterVA-5, and InSilicoVA model performance for assigning CGHR-10 CODs in the child age group (28 days to 11 years) by COD. PCCC values are sorted by the highest to lowest PCCC for each COD. Only models with PCCC values within .05 of the highest and lowest PCCC are shown. Congenital anomalies is not notable due to small sample size of 1.}\label{fig-perf-cod-child}
\end{figure}

@@

GPT-4 had the highest PCCC for 0-6 day and 7-27 day neonatal age ranges at 0.6 and 0.7 respectively, while GPT-3.5 had the lowest PCCC across all neonatal age ranges as seen in Figure \ref{fig-perf-agerange-neo}. There were large differences between minimum and maximum PCCC for 7-27 days at 0.33, and small differences for 0-6 days at 0.16. Most records were in the 0-6 day age range at 395 (83\%) records, while the rest were in the 7-27 day age range at 82 (17\%) records. Across 5 neonatal CODs (excluding congenital anomalies and other due to small sizes of 2 (\textless 1\%) and 5 (1\%) respectively), the highest PCCC was observed for GPT-4 for 3 of the 5 CODs (GPT-3.5 identical for prematurity and low birthweight), GPT-3.5 for 2 of the 5 CODs (InSilicoVA identical for stillbirth), and InSilicoVA for neonatal infections (Figure \ref{fig-perf-cod-neo}). Stillbirth had small differences between minimum and maximum PCCC at 0.07. The other 4 CODs had large differences between minimum and maximum PCCC ranging from 0.46 to 0.64. Each neonatal COD had between 23 (5\%) to 172 (38\%) records, with stillbirth (n=172, 36\%), birth asphyxia and birth trauma (n=103, 22\%), and neonatal infections (n=99, 21\%), and prematurity and low birthweight (n=73, 15\%) making 94\% of all records, while the rest are in ill-defined. GPT-4 had the highest PCCC for both female and male records at 0.62, while all other models had PCCC at or below 0.56 (see Figure \ref{fig-perf-sex-neo} in Appendix \ref{app-perf-misc}).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.7\textwidth]{fig-perf-cod-neo.pdf}
\caption{Model performance for neonatal records by COD. GPT-3.5, GPT-4, InterVA-5, and InSilicoVA model performance for assigning CGHR-10 CODs in the neonatal age group (under 28 years) by COD. PCCC values are sorted by the highest to lowest PCCC for each COD. Only models with PCCC values within .05 of the highest and lowest PCCC are shown. Congenital anomalies and other are not notable due to small sample sizes of 2 and 5 respectively.}\label{fig-perf-cod-neo}
\end{figure}
